Multi-output Bus Travel Time Prediction Using Convolutional LSTM Neural Network

### 2017-09-26 paper/paper.tex
Add figures.

\end{frontmatter}
\linenumbers
\section{Introduction}
Public transport authorities has long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) cf.\ \cite{Tcrp48}. Examples include real-time traffic information for passengers, including arrival- and departure time predictions, where studies \cite{Fan2016} show that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time.

Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g.\ demand-adaptive transit systems, and \emph{connection insurance} between public transport services. Arrival/departure time prediction is commonly approached as a specialization of travel time prediction, where the predicted travel time is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey. Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are accumulated downstream.

@@

This paper presents a multi-output, multi-time-step deep neural network for bus travel time prediction using \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers. The goal of this work is to produce precise short-term predictions (e.g.\ 0--3 hours) for link travel time, specifically for bus traffic in urban areas. The method is evaluated and compared to other popular approaches for link travel time prediction. The paper is structured in the following manner: In the next related work and literature is reviewed. \Cref{sec:convlstm} introduces Convolutional LSTM neural networks in general and in \Cref{sec:model} we present the proposed multi-output model in more details, including e.g.\ network topology, etc. \Cref{sec:experiments} introduces the dataset which the model has been evaluated on and results are presented and discussed in~\Cref{sec:results}. Finally we conclude on the work in~\Cref{sec:conclusion}.

@@

Input data is arranged in $N$ samples, each with $w$ time steps $t-w+1, \ldots, t$, and each time step with $k$ link travel times $e_1, \ldots, e_k$ cf.~\Cref{fig:data_shape}. Likewise is the output arranged with predictions of $j$ time steps ahead, $t+1, \ldots, t+j$. Thus the input is a 4D-tensor with shape $(N, w, k, 1)$, and the output a 4D-tensor with shape $(N, j, k, 1)$. It is emphasized that each prediction consists of travel time predictions for all links for the next $j$ time steps, i.e.\ multi-output, multi-time-step prediction.
\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.6\textwidth]{drawings/data_shape-crop.pdf}
  \caption{Shapes of the input and output data.}
  \label{fig:data_shape}
\end{figure}

>>>>

\end{frontmatter}
\linenumbers
\section{Introduction}
Public transport authorities has long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) \cite{Tcrp48}. Examples include real-time traffic information for passengers, e.g.\ departure boards, where studies has shown that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time \cite{Fan2016}.

Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g.\ demand-adaptive transit systems, and \emph{connection insurance} between different public transport services. Arrival/departure time prediction is commonly approached as a specialization of travel time prediction, where the predicted travel time is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey. Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are accumulated downstream.

@@

In this paper we presents a multi-output, multi-time-step deep neural network for bus travel time prediction using \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers. The goal of this work is to produce precise short-term predictions (e.g.\ 0--3 hours) for link travel time, specifically for bus traffic in urban areas. The method is evaluated and compared to other popular approaches for link travel time prediction. The paper is structured in the following manner: In the next related work and literature is reviewed. \Cref{sec:convlstm} introduces Convolutional LSTM neural networks in general and in \Cref{sec:model} we present the proposed multi-output model in more details, including e.g.\ network topology, etc. \Cref{sec:experiments} introduces the dataset which the model has been evaluated on and results are presented and discussed in~\Cref{sec:results}. Finally we conclude on the work in~\Cref{sec:conclusion}.

@@

Input data is arranged in $N$ samples, each with $w$ time steps $t-w+1, \ldots, t$, and each time step with $k$ link travel times $e_1, \ldots, e_k$ cf.~\Cref{fig:data_shape}. Likewise is the output arranged with predictions of $j$ time steps ahead, $t+1, \ldots, t+j$. Thus the input is a 4D-tensor with shape $(N, w, k, 1)$, and the output a 4D-tensor with shape $(N, j, k, 1)$. It is emphasized that each prediction consists of travel time predictions for all links for the next $j$ time steps, i.e.\ multi-output, multi-time-step prediction.
\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.8\textwidth]{drawings/data_shape-crop.pdf}
  \caption{Shapes of the input and output data.}
  \label{fig:data_shape}
\end{figure}
